Global tradeoffs between prosperity, inequality, and the environment

### 2022-07-01 paper/manuscript.Rmd
final edits for submission

---
title: "Global tradeoffs between prosperity, inequality, and the environment"
# author: |
#   | \small \texttt{\href{mailto:juan.rocha@su.se}{\nolinkurl{juan.rocha@su.se}}}
abstract: |
  Sustainability is often framed as achieving prosperity, equality, and environmental integrity: in other words, high living standards, evenly-distributed, and low-impact. But what if there is a trilemma that inhibits the joint achievement of these three key goals? We analyzed synergies and tradeoffs between these dimensions of sustainability by investigating historical data on per-capita gross national income, the Gini coefficient for income distribution, and per-capita ecological footprint. We found that no country in the world successfully solved this trilemma, but there are typologies of countries that can help identify common barriers, and strategies to overcome them. Countries typically reduce inequality and then increase prosperity and environmental impact. Some countries are trapped in a high inequality space, while prosperous countries have reduced inequality only at the expense of increasing environmental impact. Escaping this trilemma to achieve ambitious sustainability goals will likely require economic redistribution and more urgent collective action for planetary stewardship.

output:
  pdf_document:
    dev: pdf
    keep_tex: yes
    toc: no
    latex_engine: xelatex
  rtf_document:
    toc: no
csl: /Users/juanrocha/Documents/styles/nature.csl
bibliography: references_ei.bib
citecolor: blue
documentclass: article
font: Arial
fontsize: 10pt
linkcolor: blue
urlcolor: blue
header-includes:
- \usepackage{dcolumn, rotating, longtable, lineno, float, array, tabularx, inputenc}
- \setmainfont{Arial}
- \renewcommand{\abstractname}{}
- \graphicspath{{figures/}}
- \usepackage[margin=2.5cm]{geometry}
- \linenumbers
---

@@

```

Targets:

- NatComms: 5000w | 150w abstract | 4 figs or tables | Methods end 3000w | Refs
70max

- WorldDev: | 4500-10k words including all (refs, captions, tables, etc) | 300w
abstract

Current: 3957w

@@

(Aim 1000w | Current 671w)

@@

In the literature on development, such macro-level incompatibilities are not novel. For instance, it has been argued that a country cannot achieve globalization, democracy, and sovereignty in a non-volatile combination
[@RodrikD:2011]. Using resources efficiently, fairly and with maximum freedom has also been proposed as an unattainable triad [@Coyle:2011]. Additionally, researchers suggest that based on international trends, there may be persisting tradeoffs between changes in dietary improvement, disease mitigation, and environmental integrity [@Clark:2018]. In this light, we argue that another such "trilemma" that countries may face in the Anthropocene is the difficulty - or perhaps impossibility - of being prosperous and equal while achieving or maintaining low environmental impact.

For most of human history, scarcity has been a ubiquitous condition of life: material want, not sufficiency, was the lot of most people [@Maddison:2007;
@Harari:2014]. Although global prosperity has increased in recent generations, progress has been uneven. If China’s gains in poverty alleviation were removed, the United Nations’ Millennium Development Goal of halving extreme poverty between 1990 and 2015 would not have been achieved [@ki2013millennium]. And even today, approximately 10% of the world’s population still lives under the international poverty line of US\$ 1.90 dollars per day (a derisory standard that the UN itself recognizes may grossly undercount the world's impoverished[@Alston_20220]), with a much larger share barely subsisting above that threshold and aspiring for higher and more secure living standards
[@TheWor:2018]. Furthermore, in a rise of global poverty unprecedented since the end of World War II, the recent COVID-19 pandemic pushed tens-of-millions of people back beneath that already derisory line, underscoring the economic precarity that still afflicts much of the world.

@@

Indeed, environmental integrity forms the third, and arguably most fundamental, dimension of the sustainability trilemma. There is now a scientific consensus that human activities have a wide and deleterious impact on the biosphere [@Díaz:2019]. Human activities have become the primary drivers of biodiversity loss, land degradation, and climate change, among other environmental impacts
[@Díaz:2019]. At the same time, rapid population and economic growth have exceeded a number of the biophysical boundaries that delimit a “safe operating space” for humanity [@Rockström:2009; @RaworthKate:2017]. Global development in the 21st Century requires the protection and restoration of ecosystems as well as the mitigation of catastrophic climate change [@Millenniu:2005; @IPCC:2021].

Recognizing the important nexus formed by economic, social, and environmental challenges, the United Nations’ Sustainable Development Goals explicitly address prosperity, equality, and the environment [@Sachs:2015]. However, while all three of these development goals are necessary, the pursuit of any one may either help or hinder the pursuit of the other two. In this study, we ask whether there is evidence – and rationale – that all three goals can be jointly achieved, by any given country and by the world as a whole. In other words, this study attempts to answer the questions: To what extent does the sustainability trilemma exist, and if it does exist, where does it occur and why does it occur there?

@@

(Aim 800w | current 972w)

To answer these questions, we examined the historical performance of countries over the past quarter-century using publicly available data on the three dimensions of the sustainability trilemma described above (Table \ref{tab1}). We chose the country scale due to practical considerations
of data availability, but also on the recognition that many of the most
consequential policies affecting sustainability are made and implemented at this level [@Bryan:2018]. We operationalized the trilemma as the movement of countries over time in the three-dimensional space formed by the axes of prosperity, inequality, and environmental footprint  (Fig \ref{fig:trilemma}). There is a development synergy if a country moves in desired directions towards more than one goal (e.g. reducing inequality and increasing prosperity) There is a development tradeoff if positive movement towards one goal is accompanied by movement in an undesirable direction for another. Therefore, synergies are movements along
the (blue) diagonal where the achievement of two goals can be positively correlated;
while tradeoffs are represented by the (red) axis where the achievement of one
pervades the achievement of another (negative correlations). It is important to note that in this framing, the sign of the correlation changes depending on the term used, i.e. if one speaks of equality as opposed to inequality, or environmental quality as opposed to environmental footprint. In the remainder of this paper, we use inequality and environmental footprint because that is how the original data are defined and structured. Therefore, across these two domains, increasing numbers mean less desirable conditions, and negative correlations translate to synergies between the goals of prosperity, equality, and environmental integrity  (Fig \ref{fig:trilemma}).

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 4in, height = 4in]{fig1_trilemma_space}
\caption{\textbf{Trilemma space} Countries around the world face the challenge of reducing inequality and environmental impact while increasing prosperity (A). There could be synergies and trade-offs in achieving these three objectives (B-D). We explore trajectory typologies in the trilemma space. 53 trajectories from 1995 to 2017 are depicted. The trilemma is projected in each of the facets of the tridimensional cube: inequality and ecological footprint (E), inequality and prosperity (F), and ecological footprint and prosperity (G). Thresholds that delimit the desirable space are depicted as dashed lines and the colours represent negative synergies (red), positive synergies (blue), and tradeoffs (light orange) between the dimensions of the trilemma (see Methods).}
\label{fig:trilemma}
\end{figure*}
```

@@

We used these data to chart the movement of countries across the dimensions of prosperity, equality, and environmental impact over time. We further investigated whether these movements trace out typologies of countries in the trilemma space. To identify different national profiles, we calculated the magnitudes and angles of displacement of countries over time.  We then used average and standard deviation statistics to discover country clusters and performed sensitivity analyses to test the optimal number of clusters to fit. We compared ten clustering algorithms over 30 different metrics of performance, following protocols outlined by refs [@Charrad:2014tp; @Brock:2008vz]. We used principal component analysis as an ordination method for visualization, and vector-fitting to show how clusters are pulled apart in the principal component space. All data used are publicly available, and the code used for the analysis is available at: [https://github.com/juanrocha/trilemma](https://github.com/juanrocha/trilemma).

An analysis of the missing values of the publicly available data shows that a assessing the change over time in the trilemma spaces was only possible for 53 countries. The most incomplete dataset is the inequality (Figs \ref{fig:SM_gni_missing}, \ref{fig:SM_inq_missing}, \ref{fig:SM_ecofp_missing}), where for any given year at least half of the countries of the world have missing values. Based on the quality of the three datasets, we selected time series from 1995-2017 for the 53 countries with $\leqslant 30\%$ of missing values, and inputed them with a linear interpolation. This reduced dataset, although incomplete, enables the study of the dynamics of countries in the trilemmma space (Fig \ref{fig:cube_time}).

@@

No country simultaneously achieved high prosperity, low inequality, and low environmental impact. This strongly suggests that globally, there is a sustainability trilemma. At the country level, there were strong tradeoffs between inequality and environmental impact (Fig \ref{fig:trilemma}E - as the Gini indicator for income decreases, ecological footprint increases) as well as increasing prosperity and decreasing environmental impact (Fig \ref{fig:trilemma}F- per-capita GNI and ecological footprint move in the same direction). Results show synergies between increasing prosperity and decreasing inequality (Fig \ref{fig:trilemma}G), although a number of high-income countries (e.g., Sweden, New Zealand) exhibited low levels of inequality in the past, but show high variation and a trend towards greater income inequality in more recent years.

@@

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 3in]{pca_maps}
\caption{\textbf{Country typologies on the trilemma space} Principal component analysis and a senstivity analysis on clustering routines are used to identify country typologies. 3 clusters are identified when looking at the historical trajectory of countries (N=53) from 1995-2017 (A). 2 clusters are identified when aggregating over time with means and standard deviations on the trilemma dimensions regardless of data completeness. It complements the dynamic analysis with one summary statistic over time for 140 countries (B). Mean values of ecological footprint (C), inequality (D), and gross national income (E) are shown in maps for the 140 countries.}
\label{fig:pca}
\end{figure*}
```

@@

The potential mechanisms of tradeoffs and synergies remain an important research frontier. Notably, tradeoffs between inequality and prosperity (e.g., per-capita income) have been extensively studied in relation to the concept of the Kuznets Curve (KC). In this framing, the former increases as the later grows, eventually reaching an inflection point after which growing prosperity leads to lower inequality [@Kuznets:1955]. Since this relationship was first posited, there has been conflicting evidence, with some country-level studies finding strong correlations and others detecting no clear relationships [@Chancel:2020]. This has made the identification of mechanisms difficult.  Indeed, sustained inequality reduction over the past century seems to have been a consequence of contingent but systemic shocks - the world wars and the Great Depression - rather than through regularities of the global economy or government regulations  [@Pikett:2014]. Our analysis found no evidence for a KC-like relationship, instead finding that increasing inequality tended to precede growing prosperity.

In terms of the economy-environment relationship, the Environmental Kuznets Curve (EKC) has often been advanced as an axiomatic relationship: countries increase their environmental impact with growing prosperity but after reaching an inflection point of affluence, the impact of an increasingly rich society decline (Barbier 1997). As with the original KC, empirical studies of a wide range of environmental variables have come to conflicting conclusions, with some finding international evidence for this inverted-U-shaped relationship and others not (Dasgupta et al. 2002; Stern 2004). In the trilemma space, with the composite indicator of ecological footprint as the environmental indicator, we did not detect an EKC. Various mechanisms have been proposed to explain this relationship or its absence, including market liberalization, more stringent regulations, and greater freedoms for citizens to voice their concerns (Mäler 2013). But what is clear from diverse empirical studies is that the connections between prosperity and environmental integrity are mediated or confounded by other, largely social, factors.

@@

Additionally, structural inequities in the global economy are creating, or at least exacerbating, a skewed international distribution of sustainability-related development burdens. Environmental impacts can be "outsourced", as is common with affluent countries that have shifted resource-intensive production to developing countries through global supply chains [@Lambin:2011] - an important factor not captured in most EKC analyses (Mäler 2013). For instance, the US and a number of European countries have outsourced a significant portion of their carbon emissions, land use, and environmental impacts to lower-income economies through trade [@Nijdam:2005; @Peters:2006; @Weinzett:2013].  Additionally, it has been shown that international trade has significantly influenced global land use patterns, worsening natural resource exploitation and biodiversity loss[@Creutzig:2019].  This "telecoupling" of unsustainable resource use between far away places may be a mechanism that can obfuscate real trajectories in the trilemma space, undermining the ability to reach all three goals at the global level by entrapping certain countries in undesirable regions.

@@

We would like to thank the conveners and other participants of the Beijer Young Scholars workshops (2016-2018) for their insightful comments and advice as well as the Beijer Institute of Ecological Economics, at the Royal Swedish Academy of Sciences, for its institutional support.  This work was also supported by funding from Formas (grant number 2018-02318).

@@

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 6in]{FigS4}
\caption{\textbf{Trajectories examples} As shown by the experiences of Ukraine and China from 1995 to 2017, historical and structural factors influence the trajectories of countries within the “trilemma space” formed by the economic, social, and environmental dimensions of sustainability: (A) the global financial crisis; (B) the beginning of the Russo-Ukrainian conflict; (C) the accession of China into the World Trade Organization; (D) the beginning of the large-scale anti-corruption crackdown in China.}
\label{fig:S4}
\end{figure*}
```

>>>>

---
title: "Global tradeoffs between prosperity, inequality, and the environment"
author: |
    | \small Tong Wu^1,§^, Juan C. Rocha^2,3,§^, Kevin Berry^4^, Tomas Chaigneau^5^, Emilie Lindkvist^2^, Maike Hamann^6^, Jiangxiao Qiu^7^, 
    | \small Caroline Schill^2,8^, Alon Shepon^9^, Anne-Sophie Crépin^8^, and Carl Folke^2,8^
    | \small
    | \footnotesize ^1^The Natural Capital Project and Stanford Center on China's Economy and Institutions, Freeman Spogli 
    | \footnotesize Institute for International Studies, Stanford University
    | \footnotesize ^2^Stockholm Resilience Centre, Stockholm University
    | \footnotesize ^3^Future Earth
    | \footnotesize ^4^College of Business and Public Policy, University of Alaska Anchorage
    | \footnotesize ^5^Environment and Sustainability Institute, University of Exeter
    | \footnotesize ^6^Center for Complex Systems in Transition, Stellenbosch University
    | \footnotesize ^7^School of Forest Resources & Conservation, Fort Lauderdale Research and Education Center, University of Florida
    | \footnotesize ^8^Beijer Institute of Ecological Economics, The Royal Swedish Academy of Sciences
    | \footnotesize ^9^Department of Environmental Studies, The Porter School of the Environment and Earth Science, Tel Aviv University
    | \footnotesize ^§^These authors made equal contributions
#   | \small \texttt{\href{mailto:juan.rocha@su.se}{\nolinkurl{juan.rocha@su.se}}}
abstract: |

  Sustainability is often framed as achieving prosperity, equality, and environmental integrity: in other words, high living standards, evenly-distributed, and low-impact. But what if there is a trilemma that inhibits the joint achievement of these three key goals? We analyzed synergies and tradeoffs between these dimensions of sustainability by investigating historical data on per-capita gross national income, the Gini coefficient for income distribution, and per-capita ecological footprint. We found that no country in the world successfully solved this trilemma, but there are typologies of countries that can help identify common barriers, and strategies to overcome them. Countries typically reduce inequality and then increase prosperity and environmental impact. Some countries are trapped in a high inequality space, while prosperous countries have reduced inequality only at the expense of increasing environmental impact. Escaping this trilemma to achieve ambitious sustainability goals will likely require economic redistribution and more urgent collective action for planetary stewardship.
output:
  pdf_document:
    dev: pdf
    keep_tex: yes
    toc: no
    latex_engine: xelatex
  rtf_document:
    toc: no
csl: /Users/juanrocha/Documents/styles/nature.csl
bibliography: references_ei.bib
citecolor: blue
documentclass: article
font: Arial
fontsize: 10pt
linkcolor: blue
urlcolor: blue
header-includes:
- \usepackage{dcolumn, rotating, longtable, lineno, float, array, tabularx, inputenc}
- \setmainfont{Arial}
- \renewcommand{\abstractname}{}
- \graphicspath{{figures/}}
- \usepackage[margin=2.5cm]{geometry}
- \linenumbers
---

@@

## word counts
## Abstract 150
## Intro 697
## Method 957
## Results 621
## Discussion 1434
## Conclusion 267
## Total main 3976 + 150 abstract
## 
## Targets: 
# 
# - NatComms: 5000w | 150w abstract | 4 figs or tables | Methods end 3000w | Refs
# 70max
# 
# - WorldDev: | 4500-10k words including all (refs, captions, tables, etc) | 300w
# abstract
```

@@

@@

In the literature on development, such macro-level incompatibilities are not novel. For instance, it has been argued that a country cannot achieve globalization, democracy, and sovereignty in a non-volatile combination
[@RodrikD:2011]. Using resources efficiently, fairly and with maximum freedom has also been proposed as an unattainable triad [@Coyle:2011]. Additionally, research suggests that based on international trends, there may be persisting tradeoffs between changes in dietary improvement, disease mitigation, and environmental integrity [@Clark:2018]. In this light, we argue that another such "trilemma" that countries may face in the Anthropocene is the difficulty - or perhaps impossibility - of being prosperous and equal while achieving or maintaining low environmental impact.

For most of human history, scarcity has been a ubiquitous condition of life: material want, not sufficiency, was the lot of most people [@Maddison:2007;
@Harari:2014]. Although global prosperity has increased in recent generations, progress has been uneven. If China’s gains in poverty alleviation were removed, the United Nations’ Millennium Development Goal of halving extreme poverty between 1990 and 2015 would not have been achieved [@ki2013millennium]. And even today, approximately 10% of the world’s population still lives under the international poverty line of US\$ 1.90 dollars per day (a derisory standard that the UN itself recognizes may grossly undercount the world's impoverished[@Alston_20220]), with a much larger share barely subsisting above that threshold and aspiring for higher and more secure living standards
[@TheWor:2018]. Furthermore, in a rise of global poverty unprecedented since the end of World War II, the recent COVID-19 pandemic pushed tens-of-millions of people back beneath that already derisory line, underscoring the economic precarity that still afflicts much of the world [@buheji2020extent].

@@

Indeed, environmental integrity forms the third, and arguably most fundamental, dimension of the sustainability trilemma [@arrow1995economic]. Human well-being intimately depends on it and it forms the precondition for economic development. There is now a scientific consensus that human activities have a wide and deleterious impact on the biosphere, our home [@Díaz:2019; @folke2021our]. Human activities have become the primary drivers of biodiversity loss, land degradation, and climate change, among other environmental impacts
[@Díaz:2019]. Simultaneously, rapid population and economic growth have exceeded a number of the biophysical boundaries that delimit a “safe operating space” for humanity [@Rockström:2009; @RaworthKate:2017]. Global development in the 21st Century requires the protection and restoration of ecosystems as well as the mitigation of catastrophic climate change [@Millenniu:2005; @IPCC:2021].

Recognizing the important nexus formed by economic, social, and environmental challenges, the United Nations’ Sustainable Development Goals explicitly address prosperity, equality, and the environment [@Sachs:2015]. However, while all three of these development goals are necessary, the pursuit of any one may either help or hinder the pursuit of the other two. In this study, we ask whether there is evidence – and rationale – that all three goals can be jointly achieved, by any given country and by the world as a whole. In other words, this study attempts to answer the questions: To what extent does the sustainability trilemma exist, are there inherent tradeoffs between these goals, and if it does exist, where does it occur and why does it occur there?

@@

To answer these questions, we examined the historical performance of countries over the past quarter-century using publicly available data on the three dimensions of the sustainability trilemma described above (Table \ref{tab1}). We chose the country scale due to practical considerations
of data availability, but also on the recognition that many of the most
consequential policies affecting sustainability are made and implemented at this level [@Bryan:2018]. We operationalized the trilemma as the movement of countries over time in the three-dimensional space formed by the axes of prosperity, inequality, and environmental footprint  (Fig \ref{fig:trilemma}). There is a development synergy if a country moves in desired directions towards more than one goal (e.g. reducing inequality and increasing prosperity). There is a development tradeoff if positive movement towards one goal is accompanied by movement in an undesirable direction for another. Therefore, synergies are movements along
the (blue) diagonal where the achievement of two goals can be positively correlated;
while tradeoffs are represented by the (red) axis where the achievement of one
hinders the achievement of another (negative correlations). It is important to note that in this framing, the sign of the correlation changes depending on the term used, i.e. if one speaks of equality as opposed to inequality, or environmental quality as opposed to environmental footprint. In the remainder of this paper, we use inequality and environmental footprint because that is how the original data are defined and structured. Therefore, across these two domains, increasing numbers mean less desirable conditions, and negative correlations translate to synergies between the goals of prosperity, equality, and environmental integrity  (Fig \ref{fig:trilemma}).

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 4in]{fig1_trilemma_space}
\caption{\textbf{Trilemma space} Countries around the world face the challenge of reducing inequality and environmental impact while increasing prosperity (A). There could be synergies and trade-offs in achieving these three objectives (B-D). We explore trajectory typologies in the trilemma space (E-G). 53 trajectories from 1995 to 2017 are depicted. The trilemma is projected in each of the facets of the tridimensional cube: inequality and ecological footprint (E), inequality and prosperity (F), and ecological footprint and prosperity (G). Thresholds that delimit the desirable space are depicted as dashed lines and the colours represent negative synergies (red), positive synergies (blue), and tradeoffs (light orange) between the dimensions of the trilemma (see Methods).}
\label{fig:trilemma}
\end{figure*}
```

@@

We used these data to chart the movement of countries across the dimensions of prosperity, equality, and environmental impact over time. We further investigated whether these movements trace out typologies of countries in the trilemma space. To identify different national profiles, we calculated the magnitudes and angles of displacement of countries over time.  We then used average and standard deviation statistics to discover country clusters and performed sensitivity analyses to test the optimal number of clusters. We compared ten clustering algorithms over 30 different metrics of performance, following protocols outlined by refs [@Charrad:2014tp; @Brock:2008vz]. We used principal component analysis as an ordination method for visualization, and vector-fitting to show how clusters are pulled apart in the principal component space. All data used are publicly available, and the code used for the analysis is available at: [https://github.com/juanrocha/trilemma](https://github.com/juanrocha/trilemma).

An analysis of the missing values of the publicly available data shows that assessing the change over time in the trilemma spaces was only possible for 53 countries. The most incomplete dataset is the inequality (Figs \ref{fig:SM_gni_missing}, \ref{fig:SM_inq_missing}, \ref{fig:SM_ecofp_missing}), where for any given year at least half of the countries of the world have missing values. Based on the quality of the three datasets, we selected time series from 1995-2017 for the 53 countries with $\leqslant 30\%$ of missing values, and inputed them with a linear interpolation. This reduced dataset, although incomplete, enables the study of the dynamics of countries in the trilemma space (Fig \ref{fig:trilemma}).

@@

No country simultaneously achieved high prosperity, low inequality, and low environmental impact. This strongly suggests that globally, there is a sustainability trilemma. At the country level, there were strong tradeoffs between inequality and environmental impact (Fig \ref{fig:trilemma}E - as the Gini indicator for income decreases, ecological footprint increases) as well as increasing prosperity and decreasing environmental impact (Fig \ref{fig:trilemma}G- per-capita GNI and ecological footprint move in the same direction). Results show synergies between increasing prosperity and decreasing inequality (Fig \ref{fig:trilemma}G), although a number of high-income countries (e.g., Sweden, New Zealand) exhibited low levels of inequality in the past, but show high variation and a trend towards greater income inequality in more recent years.

@@

```{=tex}
\begin{figure*}[htp]
\centering
\includegraphics[width = 6.5in, height = 5in]{pca_maps}
\caption{\textbf{Country typologies on the trilemma space} Principal component analysis and a senstivity analysis on clustering routines are used to identify country typologies. Three clusters are identified when looking at the historical trajectory of countries (N=53) from 1995-2017 (A). Two clusters are identified when aggregating over time with means and standard deviations on the trilemma dimensions regardless of data completeness. It complements the dynamic analysis with one summary statistic over time for 140 countries (B). For both A and B, red arrows are vectors fitted to the data that indicate the directions along which countries and clusters differ from each other. Mean values of ecological footprint (C), inequality (D), and gross national income (E) are shown in maps for the 140 countries.}
\label{fig:pca}
\end{figure*}
```

@@

The potential mechanisms of tradeoffs and synergies remain an important research frontier. Notably, tradeoffs between inequality and prosperity (e.g., per-capita income) have been extensively studied in relation to the concept of the Kuznets Curve (KC). In this framing, the former increases as the latter grows, eventually reaching an inflection point after which growing prosperity leads to lower inequality [@Kuznets:1955]. Since this relationship was first posited, there has been conflicting evidence, with some country-level studies finding strong correlations and others detecting no clear relationships [@Chancel:2020]. This has made the identification of mechanism difficult.  Indeed, sustained inequality reduction over the past century seems to have been a consequence of contingent but systemic shocks - the world wars and the Great Depression - rather than through regularities of the global economy or government regulations  [@Pikett:2014]. Our analysis found no evidence for a KC-like relationship, instead finding that increasing inequality tended to precede growing prosperity.

In terms of the economy-environment relationship, the Environmental Kuznets Curve (EKC) has often been advanced as an axiomatic relationship: countries increase their environmental impact with growing prosperity but after reaching an inflection point of affluence, the impact of an increasingly rich society decline [@barbier1997economic]. As with the original KC, empirical studies of a wide range of environmental variables have come to conflicting conclusions, with some finding international evidence for this inverted-U-shaped relationship and others not [@dasgupta2002confronting; @stern2004rise]. In the trilemma space, with the composite indicator of ecological footprint as the environmental indicator, we did not detect an EKC, i.e. an inflection point in the prosperity-environment plane (Fig \ref{fig:cube_ensamble}). Various mechanisms have been proposed to explain this relationship or its absence, including market liberalization, more stringent regulations, and greater freedoms for citizens to voice their concerns [@MALER201325]. But what is clear from diverse empirical studies is that the connections between prosperity and environmental integrity are mediated or confounded by other, largely social, factors.

@@

Additionally, structural inequities in the global economy are creating, or at least exacerbating, a skewed international distribution of sustainability-related development burdens. Environmental impacts can be "outsourced", as is common with affluent countries that have shifted resource-intensive production to developing countries through global supply chains [@Lambin:2011] - an important factor not captured in most EKC analyses [@MALER201325]. For instance, the US and a number of European countries have outsourced a significant portion of their carbon emissions, land use, and environmental impacts to lower-income economies through trade [@Nijdam:2005; @Peters:2006; @Weinzett:2013].  Additionally, it has been shown that international trade has significantly influenced global land use patterns, worsening natural resource exploitation and biodiversity loss[@Creutzig:2019].  This "telecoupling" of unsustainable resource use between far away places may be a mechanism that can obfuscate real trajectories in the trilemma space, undermining the ability to reach all three goals at the global level by entrapping certain countries in undesirable regions.

@@

We would like to thank the conveners and other participants of the Beijer Young Scholars workshops (2016-2018) for their insightful comments and advice as well as the Beijer Institute of Ecological Economics, at the Royal Swedish Academy of Sciences, for its institutional support. This work was also supported by funding from Formas (grant number 2018-02318), and the Swedish Research Council Dnr 2018-05862 and Dnr 2018-06139.

**Author contributions:**  Conceptualization was a team effort, TW and JCR produced the first manuscript, TW, JCR and JQ analyized the data. All authors read and edit the paper.

@@

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 5in]{FigS4}
\caption{\textbf{Trajectories examples} As shown by the experiences of Ukraine and China from 1995 to 2017, historical and structural factors influence the trajectories of countries within the "trilemma space" formed by the economic, social, and environmental dimensions of sustainability: (A) the global financial crisis; (B) the beginning of the Russo-Ukrainian conflict; (C) the accession of China into the World Trade Organization; (D) the beginning of the large-scale anti-corruption crackdown in China.}
\label{fig:S4}
\end{figure*}
```
